---
title: "Portal Rodent Abundance Demo"
author: "Hao Ye"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette: 
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Portal Rodent Abundance Demo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

This vignette is a basic guide to begin exploring the Portal data.
```
longer description of the vignette
```

## Package Setup

First we load several packages:

* `portalr` is this package, which contains functions to read in the Portal data and do some basic processing.

* `dplyr` is one of the tidyverse packages for data manipulation

* `ggplot2` is one of the tidyverse packages for plotting

```{r, warning = FALSE, message = FALSE}
library(portalr)
library(dplyr)
library(ggplot2)
```

We also use some functions from other packages, but don't necessarily want to load them into our workspace. So this just checks that they are installed:

```{r}
check_for_package_installation <- function(pkg_names)
{
  message_str <- ""
  missing_pkg <- FALSE
  for(pkg_name in pkg_names)
  {
    if(!is.element(pkg_name, installed.packages()[,1]))
    {
      message_str <- paste0(message_str, "[ ] `", pkg_name, "` is not installed.\n")
      missing_pkg <- TRUE
    } else {
      message_str <- paste0(message_str, "[*] `", pkg_name, "` is installed.\n")
    }
  }
  if(missing_pkg)
  {
    stop("Some packages not found:\n", message_str)
  } else {
    message("All packages found:\n", message_str)
  }
}

check_for_package_installation(c("tidyr", 
                                 "cowplot", 
                                 "viridis"))
```


# Retrieving the Data

Note that this package does not contain the actual Portal data, which resides online in a [GitHub repository](https://github.com/weecology/PortalData).

First, we check whether we already have the data. If we don't have the data, or if the version we have isn't the most recent, we use the `download_observations` function to download the latest copy of the data.

```{r}
my_path <- "." # use current folder to store downloaded data
rodent_file <- FullPath('PortalData/Rodents/Portal_rodent.csv', my_path)

if(!file.exists(rodent_file) || # if data files are missing OR
   portalr::observations_are_new(base_folder = my_path)) # if data is not up to date
{
  download_observations(base_folder = my_path) # download from GitHub repo
}
```

The `loadData` function reads in several tables related to the rodent abundances. We won't necessarily use all of these tables, but loading this now gives us access later.

```{r}
rodent_data_all <- loadData(path = my_path)
print(summary(rodent_data_all))
```

# Rodent Abundances

The first table that we loaded (`rodent_data_all$rodent_data`) is a record of the macrofauna observed, including rodents, but also other taxa. If we just wanted to get the rodent abundance data, we could use the `abundance` function. However, going through the all of the data filtering steps helps us to get a better of what we actually end up with.

```{r}
rodent_data <- rodent_data_all$rodent_data


rodent_data <- remove_suspect_entries(rodent_data)
# filters out non-positive `period` (index of sampling trip)
# filters out NA plot values

rodent_data <- process_unknownsp(rodent_data, rodent_data_all$species_table, unknowns = TRUE)
# keep only known species codes (filters out NA)
# keep only rodents
# unidentified rodents and non-targeted rodents --> "other"

rodent_data <- remove_incomplete_censuses(rodent_data_all$trapping_table, rodent_data, incomplete = FALSE)
# filter out trips where not all plots were censused



```


We just want a basic summary of the rodent abundance data, so we can use the `abundance` function. However, by default the returned object identifies the monthly summary by `period`, which is an index, $i$, for the $i$-th sample. To convert back into an ISO date, we load up the table of sampling dates and join the two tables.

```{r}
rodent_abundance <- abundance(path = my_path) %>%
  left_join(select(newmoons_table, "period", "censusdate"),
            by = "period") %>%
  mutate(censusdate = as.Date(censusdate)) %>% # convert censusdate to `Date` type
  select(period, censusdate, everything()) # reorder columns

print(summary(rodent_abundance))
```

Let's convert the data to long format for easier facetting:
```{r}
rodent_abundance <- rodent_abundance %>%
  tidyr::gather(species, abundance, -censusdate, -period) %>%
  left_join(select(species_table, "species", "scientificname"), 
            by = "species") %>%
  rename(scientific_name = scientificname)
```

## Figure: abundance over time

```{r, fig.cap = "Monthly abundance of rodents (all plots)", fig.width = 7.5, fig.height = 10}
my_plot <- ggplot(rodent_abundance, 
                  aes(x = censusdate, y = abundance)) + 
  geom_line() + 
  facet_wrap(~scientific_name, scales = "free_y", ncol = 3) + 
  xlab("Date") + 
  ylab("Abundance") + 
  scale_x_date(breaks = seq(as.Date("1977-01-01"), to = as.Date("2018-01-01"), "+5 years"), date_labels = "%Y") + 
  cowplot::theme_cowplot() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

print(my_plot)
```

## Next Steps

Our next steps would likely be to dig deeper into the rodent abundances for different treatments, but first we want to know what the different treatments look like, so let's revisit the abundances later.

# Plot Treatments

A description of the experimental design and treatments can be found in this [Readme file](https://github.com/weecology/PortalData/tree/master/SiteandMethods) in the PortalDate repo.

For now, we are just going to use the [`Portal_plots`](https://github.com/weecology/PortalData/blob/master/SiteandMethods/Portal_plots.csv) table file to look at how the treatments for individual plots have changed over time. Note that this file is already loaded in as the `plots_table` from the `loadData` function we ran previously.

```{r}
print(summary(plots_table))
```

We want a proper date variable as well as converting `plot` into a factor:

```{r}
plot_treatments <- plots_table %>%
  mutate(iso_date = as.Date(paste0(year, "-", month, "-", "01")), 
         plot = as.factor(plot)) %>%
  select(iso_date, plot, treatment)
```



## Figure: plot treatments over time

```{r, fig.cap = "Treatments of Plots by Date", fig.width = 7, fig.height = 10}
my_plot <- ggplot(plot_treatments, 
                  aes(x = iso_date, y = treatment, color = treatment)) +
  geom_point(shape = 20) + 
  geom_vline(aes(xintercept = as.Date("1977-10-01")), linetype = 2) + 
  geom_vline(aes(xintercept = as.Date("1988-01-01")), linetype = 2) + 
  geom_vline(aes(xintercept = as.Date("2005-01-01")), linetype = 2) + 
  geom_vline(aes(xintercept = as.Date("2015-04-01")), linetype = 2) + 
  facet_wrap(~plot, ncol = 4) + 
  xlab("Date") + 
  ylab("Treatment") + 
  scale_color_manual(values = viridis::viridis(4)) + 
  scale_x_date(breaks = seq(as.Date("1977-01-01"), to = as.Date("2018-01-01"), "+5 years"), date_labels = "%Y") + 
  cowplot::theme_cowplot() + 
  guides(color = FALSE) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

print(my_plot)
```



